---
title: 'Data Science as a Field: COVID 19'
author: 
date: "2025-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r} 
#Required libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(prophet)
```

```{r}
# Import the data
url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names = c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv")
urls = str_c(url_in, file_names)
global_conf = read.csv(urls[3])
US_conf = read.csv(urls[1])
global_deaths = read.csv(urls[4])
US_deaths = read.csv(urls[2])
uid_url = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
Pop = read.csv(uid_url)
```

## Data Preparation

### Reshaping, Conversion, and Renaming of the Global Dataset

```{r}
# Remove Lat and Long, change the dates to single column with values as cases
global_conf = global_conf %>%
 select(-Lat, -Long) %>%
 pivot_longer(
 cols = starts_with("X"),
 names_to = "Date",
 values_to = "Cases"
 ) %>%
 mutate(
 Date = sub("^X", "", Date),
 Date = as.Date(Date, format = "%m.%d.%y")
 )
```

```{r}
# Remove Lat and Long, change the dates to single column with values as cases
global_deaths = global_deaths %>%
  select(-Lat, -Long) %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Date",
    values_to = "Cases"  
  ) %>%
  mutate(
    Date = sub("^X", "", Date),
    Date = as.Date(Date, format = "%m.%d.%y")
  )
```

```{r}
# Join the global deaths and confirmed cases
global_combined <- full_join(
  global_conf,
  global_deaths,
  by = c("Province.State", "Country.Region", "Date"),
  suffix = c("_confirmed", "_deaths")
)
```

```{r}
# Remove any cases that are 0
global_combined = global_combined %>% filter(Cases_confirmed > 0)
```

### Reshaping, Conversion, and Renaming of the US Dataset
```{r}
# Change columns to combined, remove X in front of date, change dates to one column with cases for the values
US_conf = US_conf %>%
  pivot_longer(
    cols = -(UID:Combined_Key),  
    names_to = "date",
    values_to = "case"
  ) %>%
  mutate(
    date = sub("^X", "", date),  
    date = mdy(date)             
  ) %>%
  select(Admin2, Province_State, Country_Region, Combined_Key, date, case)

```

```{r}
# Change columns to combined, remove X in front of date, change dates to one column with cases for the values
US_deaths = US_deaths %>%
  pivot_longer(
    cols = -c(UID:Combined_Key, Population),  
    names_to = "date",
    values_to = "case"
  ) %>%
  mutate(
    date = sub("^X", "", date),
    date = as.Date(date, format = "%m.%d.%y")
  ) %>%
  select(Admin2, Province_State, Country_Region, Combined_Key, Population, date, case)
```

```{r}
# Combined US deaths and cases
US_combined = full_join(
  US_conf,
  US_deaths,
  by = c("Admin2", "Province_State", "Country_Region", "Combined_Key", "date"),
  suffix = c("_confirmed", "_deaths")
) %>%
  rename(
    Confirmed = case_confirmed,
    Deaths = case_deaths
  ) %>%
  select(Admin2, Province_State, Country_Region, Combined_Key, date, Population, Confirmed, Deaths)

```

```{r}
# Create the Combined_Key column and rename the columns
global_combined = global_combined %>%
  unite(Combined_Key, Province.State, Country.Region, sep = "_", remove = FALSE) %>%
  rename(Province_State = Province.State, Country_Region = Country.Region)
```

```{r}
# Renaming columns to match for the population join
global_combined = global_combined %>%
  rename(Cases = Cases_confirmed, Deaths = Cases_deaths)

US_combined = US_combined %>%
  rename(Cases = Confirmed, Date= date)
```

```{r}
# Add population to global variable
global_combined = global_combined %>%
  left_join(Pop %>% select(Province_State, Country_Region, Population), 
            by = c("Province_State", "Country_Region")) %>%
  select(Province_State, Country_Region, Date, Cases, Deaths, Population, Combined_Key)
```

## Data Analysis, Visualization, and Modeling

```{r}
# US by state
US_by_state = US_combined %>%
  group_by(Province_State, Date) %>%
  summarize(
    Cases = sum(Cases, na.rm = TRUE),
    Deaths = sum(Deaths, na.rm = TRUE),
    Population = sum(Population, na.rm = TRUE),  #
    .groups = "drop"  
  )
US_by_state = US_by_state %>%
  mutate(deaths_per_mill = Deaths * 1e6 / Population)

summary(US_by_state)

```

```{r}
# Creating US totals
US_totals = US_by_state %>%
  group_by(Date) %>%  
  summarize(
    Cases = sum(Cases, na.rm = TRUE),
    Deaths = sum(Deaths, na.rm = TRUE),
    Population = sum(Population, na.rm = TRUE),
    .groups = "drop"  
  ) %>%
  mutate(deaths_per_mill = Deaths * 1e6 / Population) %>%  
  select(Date, Cases, Deaths, Population, deaths_per_mill)  
summary(US_totals) 
```
```{r}
# Plot of US total cases and deaths
US_totals %>%
  filter(Cases > 0) %>%  
  ggplot(aes(x = Date)) +  
  geom_line(aes(y = Cases, color = "Cases")) +
  geom_point(aes(y = Cases, color = "Cases")) +
  geom_line(aes(y = Deaths, color = "Deaths")) +
  geom_point(aes(y = Deaths, color = "Deaths")) +
  scale_y_log10() +  
  scale_color_manual(values = c("Cases" = "blue", "Deaths" = "red")) +  
  theme_minimal() +  
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  labs(
    title = "COVID-19 in the US",
    y = "Total Cases & Deaths",
    x = "Date",
    color = "Legend"
  )
```
```{r}
# Texas cases and deaths
US_by_state %>%
  filter(Province_State == "Texas") %>%  
  filter(Cases > 0) %>%  
  ggplot(aes(x = Date)) +  
  geom_line(aes(y = Cases, color = "Cases")) +
  geom_point(aes(y = Cases, color = "Cases")) +
  geom_line(aes(y = Deaths, color = "Deaths")) +
  geom_point(aes(y = Deaths, color = "Deaths")) +
  scale_y_log10() +  
  scale_color_manual(values = c("Cases" = "blue", "Deaths" = "red")) +  
  theme_minimal() +  
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  labs(
    title = "COVID-19 in Texas",  
    y = "Total Cases & Deaths",
    x = "Date",
    color = "Legend"
  )
```

```{r}
# max by date
cat("Max Date:", format(max(US_totals$Date), "%Y-%m-%d"), "\n")

# max deaths
cat("Max Deaths:", (max(US_totals$Deaths)), "\n")

```
```{r}
# evaluate if the deaths are plateauing

US_by_state <- US_by_state %>%
  arrange(Province_State, Date) %>% 
  mutate(
    new_cases = Cases - lag(Cases, default = 0),
    new_deaths = Deaths - lag(Deaths, default = 0)
  )

US_totals <- US_totals %>%
  mutate(
    new_cases = Cases - lag(Cases, default = 0),
    new_deaths = Deaths - lag(Deaths, default = 0)
  )

tail(US_totals %>% select(new_cases, new_deaths, everything()))
```
```{r}
# Plot the new cases and new deaths

US_totals %>%
  ggplot(aes(x = Date)) +  
  geom_line(aes(y = new_cases, color = "New Cases")) +
  geom_point(aes(y = new_cases, color = "New Cases")) +
  geom_line(aes(y = new_deaths, color = "New Deaths")) +
  geom_point(aes(y = new_deaths, color = "New Deaths")) +
  scale_y_log10() +  
  scale_color_manual(values = c("New Cases" = "blue", "New Deaths" = "red")) +  
  theme_minimal() + 
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  labs(
    title = "COVID-19 in the US",
    y = "Total Cases & Deaths",
    x = "Date",
    color = "Legend"
  )
```

```{r}
# Texas graph of new deaths and cases

state = "Texas"  

US_by_state %>%
  filter(Province_State == state) %>%  
  filter(new_cases > 0) %>%  
  ggplot(aes(x = Date)) +  
  geom_line(aes(y = new_cases, color = "New Cases")) +
  geom_point(aes(y = new_cases, color = "New Cases")) +
  geom_line(aes(y = new_deaths, color = "New Deaths")) +
  geom_point(aes(y = new_deaths, color = "New Deaths")) +
  scale_y_log10() +  
  scale_color_manual(values = c("New Cases" = "blue", "New Deaths" = "red")) +  
  theme_minimal() +  
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  labs(
    title = "COVID-19 in Texas",  
    y = "New Cases & Deaths",
    x = "Date",
    color = "Legend"
  )
```

```{r}
# Summarize state level totals
US_state_totals = US_by_state %>%
  group_by(Province_State) %>%
  summarize(
    deaths = max(Deaths, na.rm = TRUE),  
    cases = max(Cases, na.rm = TRUE),
    population = max(Population, na.rm = TRUE),
    cases_per_thou = 1000 * cases / population,
    deaths_per_thou = 1000 * deaths / population,
    .groups = "drop"  
  ) %>%
  filter(cases > 0, population > 0)

# Get the 10 states with the lowest deaths per thousand
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%  
  select(Province_State, deaths_per_thou, cases_per_thou, cases, deaths, population)

```

```{r}
# Top 10 worse states
US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%  
  select(Province_State, deaths_per_thou, cases_per_thou, cases, deaths, population)

```

```{r}
# Linear model predicting deaths per 1000 based on cases per 1000
mod = lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)  
summary(mod)  
```

```{r}
# Deaths per 1000 by population
mod2 = lm(deaths_per_thou ~ cases_per_thou + population, data = US_state_totals)
summary(mod2)
```

```{r}
# Analysis of variance deaths vs cases per thousand and population
anova(mod, mod2)
```

```{r}
# State with the minimum cases per 1000
US_state_totals %>%
  slice_min(cases_per_thou, n = 1)
```

```{r}
# State with the maximum cases per 1000
US_state_totals %>%
  slice_max(cases_per_thou, n = 1)  

```
```{r}
# Create a sequence from 1 to 151
x_grid <- seq(1, 151)

# Create a new tibble for cases_per_thou
new_df = tibble(cases_per_thou = x_grid)

# Add predicted values from the regression model
US_state_totals = US_state_totals %>%
  mutate(pred = predict(mod, newdata = US_state_totals))  

head(US_state_totals)
```

```{r}
library(dplyr)

US_total_w_pred = US_state_totals %>%
  mutate(
    pred = predict(mod, newdata = US_state_totals),
    pred2 = predict(mod2, newdata = US_state_totals),
    std_ratio = ((deaths_per_thou / cases_per_thou) - 
                (mean(deaths_per_thou, na.rm = TRUE) / mean(cases_per_thou, na.rm = TRUE))) / 
                sd(deaths_per_thou / cases_per_thou, na.rm = TRUE)  
  )

US_total_w_pred
```
```{r}
# US total with the 2 prediction models
US_total_w_pred %>%
  ggplot(aes(x = cases_per_thou)) +
  geom_point(aes(y = deaths_per_thou), color = "blue", alpha = 0.6, size = 3) +  
  geom_point(aes(y = pred), color = "red", alpha = 0.6, size = 3) +  
  geom_point(aes(y = pred2), color = "green", alpha = 0.6, size = 3) +  
  labs(
    title = "COVID-19 Deaths per Thousand vs Cases per Thousand",
    x = "Cases per Thousand",
    y = "Deaths per Thousand",
    caption = "Blue = Actual, Red = Model 1 Prediction, Green = Model 2 Prediction"
  ) +
  theme_minimal()
```
```{r}
# Clear memory
gc()
```
```{r}
# Timeseries Forcasting


# Load and prepare data
US_combined = US_combined %>%
    mutate(Date = as.Date(Date))

# Aggregate cases for all states (sum across states for each date)
us_aggregated = US_combined %>%
    group_by(Date) %>%
    summarise(y = sum(Cases, na.rm = TRUE)) %>%
    ungroup()

# Rename columns for Prophet
colnames(us_aggregated) = c("ds", "y")

# Convert cumulative cases to daily new cases
us_aggregated = us_aggregated %>%
    arrange(ds) %>%
    mutate(y = y - lag(y, default = first(y))) %>%  
    mutate(y = ifelse(y < 0, 0, y))  

# Apply log transformation
us_aggregated = us_aggregated %>%
    mutate(y = log1p(y))  

# Train Prophet model
model = prophet(us_aggregated, 
                 weekly.seasonality = TRUE, 
                 yearly.seasonality = TRUE)

# Generate future predictions
future_dates = make_future_dataframe(model, periods = 30)
forecast = predict(model, future_dates)

# Convert back from log scale
forecast = forecast %>%
    mutate(yhat = expm1(yhat),
           yhat_lower = expm1(yhat_lower),
           yhat_upper = expm1(yhat_upper))

# Plot the forecast
plot(model, forecast)

# Plot seasonality components (trend, weekly, yearly)
prophet_plot_components(model, forecast)
```
#### First Image (Seasonality Components). Second Image (Actual vs Predicted Cases): Black Dots: Actual COVID-19 daily cases.Blue Line: Forecasted values with confidence intervals. ds = date stamp. y = responsive variable (daily covid cases).

## Conclusions, Biases

### COVID-19 was considered a serious global threat, and understanding its spread is crucial for epotential future infectious issues. This analysis examines how population size and case rates per thousand influence deaths per thousand across U.S. states. Using linear regression models, I generated state-level death rate predictions. However, the models could be significantly improved with additional predictors such as lockdown, masking, testing rates, vaccinations, and temperature.

### Timeseries forecasting for the US states: I created a timeseries forecast as an aggregate of all the states. This hows a rise in cases in early 2020, peaks in 2021–2022, and then stabilizes. Cases increase early in the week (Monday–Wednesday) and drop on weekends (Saturday). eaks appear around April and December, which might align with holiday surges or seasonal COVID-19 waves. Actual vs Predicted Cases-> This model captures major COVID-19 waves (2021 & 2022 spikes). 

### A key source of bias in the data stems from inconsistencies in state-level reporting of cases and deaths, influenced by infrastructure limitations and reporting delays. Testing rates also introduce bias, as underreporting of asymptomatic or untested cases may lead to an underestimation of true infection rates. Additionally, my focus on U.S. data introduces personal bias, as a broader global dataset might yield different insights. The choice of an alpha level of 0.05 could also impact statistical conclusions.

### Despite these limitations, the data indicates a decline in COVID-19 cases, reinforcing the importance of continued safety measures and vaccine distribution to mitigate the pandemic’s impact.

